# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r cache = TRUE}  
data <- read.csv("activity.csv") # make sure the data file is in your working directory
data$date <- as.character(data$date)  # transform the data type of variable "date" and "interval"
data$date <- as.Date(data$date)
data$interval <- as.factor(data$interval)
```

## What is mean total number of steps taken per day?
```{r cache = TRUE}
dates <- seq(from = min(data$date), to = max(data$date), by = 1)
# Calculate the total number of steps taken per day
steps.per.day <- data.frame()
for(date in dates) {
    steps.per.day <- rbind(steps.per.day, data.frame(steps = sum(data[data$date == date, ]$steps, na.rm = TRUE)))
}
# Make a histogram of the total number of steps taken each day
library("lattice")
histogram(~ steps, data = steps.per.day, xlab = "Steps taken per day")
mean <- mean(steps.per.day$steps) # calculate the mean
median <- median(steps.per.day$steps) # calculate the median
```
The mean and median total number of steps taken per day is `r mean` and `r median`.

## What is the average daily activity pattern?
```{r cache = TRUE}
library(reshape2)
remove(mean) # remove the variable "mean" in order to use dcast function
data.melting <- melt(data, id = c("interval"), measure.vars = c("steps"), na.rm = TRUE)
data.casting <- dcast(data.melting, interval ~ variable , mean)
xyplot(steps ~ interval, data = data.casting, type = 'l', ylab = "Average number of steps", xlab = "Interval")
# Get the interval containing the maximum number of steps
interval.max <- data.casting[data.casting$steps == max(data.casting$steps), ]$interval
```
The interval `r interval.max` contains the maximum number of steps, which is about `r max(data.casting$steps)`

## Imputing missing values
```{r cache = TRUE}
counts.missing <- sum(is.na(data)) # calculate the total number of missing values
index.missing <- which(is.na(data))
# Create a new data set and fill in all of the missing values with the mean for that 5-minute interval
data.new <- data
for(index in index.missing) {
    interval <- data.new[index, ]$interval
    data.new[index, ]$steps <- data.casting[data.casting$interval == interval, ]$steps
}
# Make a histogram of the total number of steps taken each day using the new dataset
steps.per.day.new <- data.frame()
for(date in dates) {
    steps.per.day.new <- rbind(steps.per.day.new, data.frame(steps = sum(data.new[data.new$date == date, ]$steps)))
    }
histogram(~ steps, data = steps.per.day.new, xlab = "Steps taken per day")
mean.new <- mean(steps.per.day.new$steps) # calculate the new mean
median.new <- median(steps.per.day.new$steps) # calculate the new median
```
The total number of missing values in the dataset is `r counts.missing`.  
The new mean and median total number of steps taken per day is `r mean.new` and `r median.new`.  
We can see that the new mean and median values are both larger for we replace the missing values with the positive average 5-minite interval value.

## The differences between weekday and weekend activity pattern
```{r cache = TRUE}
# Create a new factor variable called "weekday" to indicate whether a given date is a weekday or weekend day.
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
weekday <- numeric()
for(i in 1:nrow(data.new)) {
    weekday <- c(weekday, ifelse(weekdays(data.new[i, ]$date) %in% weekdays, 1, 0))
    }
weekday <- as.factor(weekday)
levels(weekday) <- c("weekend", "weekday")
# Add this factor varialbe to the dataset.
data.new <- cbind(data.new, weekday)
# Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.
remove(mean) # remove the variable "mean" in order to use dcast function
data.new.melting <- melt(data.new, id = c("interval", "weekday"), measure.vars = c("steps"))
data.new.casting <- dcast(data.new.melting, interval + weekday ~ variable , mean)
xyplot(steps ~ interval|weekday, data = data.new.casting, type = 'l', ylab = "Average number of steps", xlab = "Interval", layout = c(1, 2))
```
